---
title: "Investigating the Relationship Between Ticket Prices & Seats Sold for Broadway’s The Lion King & Hamilton"
author: Reem Alasadi
thanks: "Code and data are available at: https://github.com/alasadir/broadway"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "A linear regression model is conducted to analyze the relationship between top and average ticket prices and the number of seats sold for the Broadway shows 'The Lion King' and 'Hamilton'. Results show that these variables are correlated, but do not explain what affects ticket prices. This suggests a need to explore other affecting variables."
keywords: 
- broadway
- the lion king
- hamilton
- multiple linear regression
- ticket prices
- seats sold
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(kableExtra)
library(broom)
library(huxtable)
library(tidymodels)
library(rsample)

grosses <- read_csv(here::here("inputs/data/grosses.csv"),
                     guess_max = 40000)

hamilton_and_lion_king <- 
  grosses %>% 
  filter(show %in% c("Hamilton", "The Lion King"))

```

\newpage

# Introduction

I am a modest musical fan. I know a good number of musicals thanks to Disney and my theater-fanatic friend from high school, Hadeel. My 2 favorite theatrical musicals are The Lion King and Hamilton. While browsing Alex Cookson’s blog [@alex], his take on the most successful Broadway show of all time grabbed my attention. He analyzes box office returns to determine the highest-grossing show. Which got me wondering why are ticket prices set the way they are? How are ticket prices affected by seats sold? 

This project builds on Alex’s work by creating a multiple regression model to try to understand how ticket prices and seats sold for The Lion King and Hamilton are correlated. The linear regression model should quantify the strength of the relationship between the ticket price and the seats sold. If a relationship is found, it could help forecast future ticket prices.

Two models are run, one with the top ticket price as the dependent (response) variable, and another with the average ticket price as the dependent variable. It was found that top ticket price, average ticket price and the number of seats sold are correlated. The results, however, do not suggest that these variables can predict or explain ticket prices.

The [Method and Data](#M&D) section will give information about the dataset and some exploratory analysis. The [Model](#model) section will cover how the multiple linear regression was conducted, and the [Results](#results) section will report the results of the model. The [Discussion](#discusssion) section will discuss the implications of the results, and the limitations of the model.


# Ethical Considerations

While I believe that what was done with The Lion King franchise is phenomenal, I feel responsible to give a nod to Osamu Tezuka's anime series 'Kimba the White Lion' (1965) and the 'Jungle Emperor' (1950) manga, whose work influenced The Lion King, despite the controversy. There is plenty of information about the controversy online, but I think Pete Keely wrote an encompassing summary on the issue on [*The Hollywood Reporter*][1] [@keeley_2020].

[1]:https://www.hollywoodreporter.com/heat-vision/lion-king-kimba-white-lion-does-disney-need-come-clean-1225822

# Method and Data: {#M&D}

## Method: 

The data will be analyzed using `R` [@citeR], a statistical programming language, and packages that allow easier data manipulation, like: `tidyverse` [@tidy], `lubridate` [@lubr], `knitr` [@knitr], and `rsample`. Tables will be formatted using `KableExtra` [@KablExtra] and `huxtable` [@hux]. Data will be visualized using the `tidymodels` package [@tidy] that includes `ggplot2` package [@gg].

The multiple linear regression model will compare the correlation between top_ticket, avg_ticket and seats_sold variables for The Lion King and Hamilton. Top ticket and average ticket prices will alternate as dependent (response) variables in the model.

The code for visualizing confidence intervals was taken from `rsample`'s website [@confidenceInt].

## Dataset description:

The dataset includes information about many Broadway shows from the very first week they started showing in theaters on Broadway, New York until the latest performance in March 2020, as all shows were stopped around that time because of the COVID-19 pandemic. The dataset includes information like the number of weeks the show has been running, the overall gross, the theater the show was performed at, the number of seats sold, average and top ticket prices, etc. The information has been scraped off of the Playbill website, which uses data from The Broadway League's official website. The Broadway League is the official entity that oversees all theatrical shows on Broadway.

The original dataset contains 47,524 observations and 14 variables. Since the scope of the project will be on The Lion King and Hamilton shows only, the dataset has been filtered and now has 1,407 observations with 3 N/A values and 14 variables. Here's a preview of how it looks like:

```{r previewofdataset, fig.height=8, fig.width=1, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
head(hamilton_and_lion_king, 1) %>%
  knitr::kable(caption = "", digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE, latex_options = "scale_down")
```


Below is a list of the variable labels:

```{r preview, echo=FALSE}
colnames(hamilton_and_lion_king)
```

Below are summary statistics:

```{r, results='asis'}
summary(hamilton_and_lion_king) %>%
  knitr::kable(caption = "Dataset Summary", digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE)
```


```{r message=FALSE, include=FALSE}
sum(is.na(hamilton_and_lion_king))
```


## Variables used in the model:

The variables that will be analyzed are as follows. Variable descriptions are imported from Alex’s repo [@alex]:

| variable | class | description |
|:-:|:-:|:-:|
| show | character | Name of the show, i.e. The 'Lion King' and 'Hamilton' |
| avg_ticket_price | double | Average price of tickets sold |
| top_ticket_price | double | Highest price of tickets sold |
| seats_sold | double | Total seats sold for all performances and previews |

The only change made to a variable is the data-type for 'show'; it is changed from 'character' to 'factor' as it contains binary data. This change is necessary for the model to recognize categorical variables.


## Data Exploration

The Lion King is considered to be one of the longest and most successful Broadway shows in history [@lk]. It has been running since October 1997 (over 20 years) and grossed over \$1.6 billion in performances on Broadway alone (not counting performances done on tour). Hamilton, on the other hand, first debuted in October 2015 but has gained much success. In fact, in the 6 years of running, Hamilton grossed over $600 million. That is equivalent to what The Lion King made in about 12 years. Figure \@ref[fig:grosses] exhibits the grosses of both shows over the years:

```{r grosses, fig.cap="Grosses over the years", echo=FALSE}
hamilton_and_lion_king %>% 
  ggplot(aes(x = week_ending, y = weekly_gross/1000000)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Weekly Gross (in millions)") +
  xlab("") +
  theme_minimal()
```

\newpage

Figure \@ref(fig:performances) exhibits that both shows maintained 8 performances per week most of the time:

```{r performances, fig.cap="Performances per week", echo = FALSE}
hamilton_and_lion_king %>% 
  mutate(performances = as.integer(performances)) %>% 
  ggplot(aes(x = week_ending, y = performances)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Number of Performances") +
  xlab("") +
  theme_minimal()
```

\newpage

As shown in figure \@ref(fig:aveticketsvsseatssold), Hamilton consistently sold 10,000 seats per week at average ticket price, and The Lion King sold between 13 - 15 thousand seats:

```{r aveticketsvsseatssold, fig.cap="Seats sold per week versus average ticket prices", echo = FALSE}
hamilton_and_lion_king %>% 
  ggplot(aes(x = avg_ticket_price, y = seats_sold)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Seats Sold per Week") +
  xlab("Average Ticket Price (USD)") +
  theme_minimal()

```

\newpage

So the number of seats sold for Hamilton is less than The Lion King’s, but what about the theater’s capacity?

```{r aveticketsvspctcapacity, fig.cap="Percentage of theater capacity sold versus average ticket price", echo = FALSE}
hamilton_and_lion_king %>% 
  ggplot(aes(x = avg_ticket_price, y = pct_capacity*100)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Theatre Capacity (%)") +
  xlab("Average Ticket Price (USD)") +
  theme_minimal()
```


Figure \@ref(fig:aveticketsvspctcapacity) shows that when the average ticket price for The Lion King was between about \$50 - \$200, the theater was at 100% capacity, meaning it sold out. The demand for Hamilton was stronger; average tickets sold out when they ranged between \$150 - \$350. Some shows even had over 100% capacity. These are instances of standing room tickets.

\newpage

Time to explore the ticket prices.


```{r aveticketprice, echo= FALSE, message=FALSE, fig.cap="Average ticket prices each week"}
hamilton_and_lion_king %>% 
  ggplot(aes(x = week_ending, y = avg_ticket_price)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Average Ticket Price (USD)") +
  xlab("") +
  theme_minimal()
```


It is evident in figure \@ref(fig:aveticketprice) that Hamilton tickets are way more expensive than The Lion King’s. Hamilton may have started close to the lowest ticket price of The Lion King in 2015, but the average ticket price for Hamilton doubled within the first year and exceeded The Lion King's within the second year. While the average ticket price for The Lion King could reach up to a maximum of \$200, Hamilton’s _started_ at \$250. 

\newpage

Similarly, the top ticket price for Hamilton starts at the maximum of The Lion King's tickets. It is almost double the price. Figure \@ref(fig:topticketprice) shows the top ticket price trend:

```{r topticketprice, echo=FALSE, fig.cap="Top ticket prices per week", message=FALSE, warning=FALSE}
hamilton_and_lion_king %>% 
  ggplot(aes(x = week_ending, y = top_ticket_price)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Top Ticket Price (USD)") +
  xlab("") +
  theme_minimal()
```

\newpage

To put things into perspective, figure \@ref(fig:aveticketsvstopticketpricebyyears) compares top ticket price versus average ticket price for both shows over the years. The Lion King's ticket prices had incremental increases in the last 5 years. However, Hamilton's ticket prices grew exponentially over the first 2 years, and gradually for the remaining years.

```{r aveticketsvstopticketpricebyyears, fig.cap="Top ticket prices versus average ticket prices per year", echo = FALSE}
hamilton_and_lion_king %>% 
  mutate(year_of_show = lubridate::year(week_ending)) %>%
  filter(year_of_show %in% c("2015", "2016", "2017", "2018", "2019", "2020")) %>%
  ggplot(aes(x = avg_ticket_price, y = top_ticket_price)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Top Ticket Price (USD)") +
  xlab("Average Ticket Price (USD)") +
  theme_minimal() +
  facet_wrap(vars(year_of_show))

```

\newpage

# Model

Linear regression is one of the statistical analysis tools that shows the relationship between two or more variables. It uses data to predict trends. As mentioned earlier, the linear regression model is used to quantify the strength of the relationship between the ticket prices and the seats sold. It can also help forecast future ticket prices. 

## Alternative Models

An Autoregressive Integrated Moving Average (ARIMA) model could be also used to understand current data and forecast future ticket prices. It relies on historical data to make predictions. I would have loved to have conducted both multiple linear regression and ARIMA and compared the results, but ARIMA uses time-series data and that is beyond my humble expertise in R. I tried.


## Model Specifics

Two regression models will be created; one uses the top ticket price as the dependent variable, and the other uses the average ticket price. To distinguish between both models, the top ticket price model will be referred to as the top_model, and the average ticket price model will be referred to as avg_model.

**Top_model specifics:**
- Dependent variable: Top ticket price
- Independent variables: Average ticket price, seats sold, show
- Mathematical notation:
$$
Top Ticket Price = Intercept (b) + m (average ticket price) + m (seats sold) + m (show) + residual error
$$

**Avg_model specifics:**

- Dependent variable: Average ticket price
- Independent variables: Top ticket price, seats sold, show
- Mathematical notation:
$$
Average Ticket Price = Intercept (b) + m (top ticket price) + m (seats sold) + m (show) + residual error
$$


## Hypothesis

**H0:** There is no statistically significant relationship between ticket prices and seats sold at a show

**H1:** There is a statistically significant relationship between ticket prices and seats sold at a show


# Results {#results}


```{r , echo=FALSE, include=FALSE}

#since 'show' contains binary data, Lion King & Hamilton, it is considered as categorical data and should be converted to a factor

hamilton_and_lion_king <-
hamilton_and_lion_king %>%
mutate(show = as_factor(show))
```

```{r echo=FALSE, message=FALSE, include=FALSE}
#checking the data type for 'show' to make sure changes were made

class(hamilton_and_lion_king$show)
```

```{r echo=FALSE, message=FALSE}
#checking the order of the shows within the column. This will help determine which show will be taken as a dependent variable in the lm() model

levels(hamilton_and_lion_king$show)
```

```{r model variables, message=FALSE, include=FALSE}
model_data <- hamilton_and_lion_king %>%
  select(top_ticket_price,
         avg_ticket_price,
         seats_sold,
         show)
summary(model_data)
```

```{r message=FALSE, include=FALSE}
dim(model_data)
```

```{r include=FALSE}
plot(model_data)
```

## Top Ticket Price Model

```{r model top ticket, echo=FALSE, message=FALSE}
top_model <- lm(formula = top_ticket_price ~
                avg_ticket_price +
                seats_sold + 
                show,  
             data = hamilton_and_lion_king)
summary(top_model)
```

$$
Top Ticket Price = 308.31 + 1.7 (average ticket price) - 0.02 (seats sold) + 223.81 (show) + 64.56 
$$

The results show that there is a statistically significant relationship between the top ticket price and seats sold at a show. Therefore the null hypothesis is rejected.

|  | Value | Interpretation |
|:-:|:-:|:-:|
| R-Squared | 92.02% | The R-squared value shows that about 92.02% of the variation in the top ticket price can be explained by the top model. |
| Intercept | 308.32 | We would expect an average of \$308.32 for the top ticket price for The Lion King, when the average ticket price is \$0 and 0 seats are sold. |
| Average ticket | 1.74 | While holding the other variables constant, for each \$1 dollar difference in the average ticket price for The Lion King, an average of $1.74 difference is expected for top ticket price |
| Seats sold | -0.02 | While holding the other variables constant, for each seat sold for The Lion King, the top ticket price is expected to decrease by an average of $0.02. |
| Hamilton show | 223.82 | Compared to The Lion King, we would expect that the top ticket price for Hamilton to be $233.82 more expensive, while holding other variables constant. |


### Plotting Top Model

```{r plotting top_model, echo=FALSE, message=FALSE, warning=FALSE}
hamilton_and_lion_king %>%
  ggplot(aes(x = avg_ticket_price, seats_sold, y = top_ticket_price, color = show)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(x = "Average Ticket Price (USD) + Seats Sold",
       y = "Top Ticket Price (USD)",
       color = "Show") +
  theme_classic()
```


### Plotting Top Model Residuals

```{r residuals, message=FALSE, include=FALSE}
broom::augment(top_model)
```

```{r plotting residuals, echo=FALSE}
ggplot(top_model, 
       aes(x = .resid)) + 
  geom_histogram(binwidth = 1) +
  theme_classic() +
  labs(y = "Number of occurrences",
       x = "Residuals")
```

### Top Model Confidence Intervals [

```{r top model diagnostics, include=FALSE}
glance(top_model)
```

```{r top model coefficient estimates, include=FALSE}
tidy(top_model)
```

```{r include=FALSE}
confint(top_model)
```

```{r top model confidence interval prep, echo=FALSE}
set.seed(1407)

top_model_intervals <- 
  reg_intervals(top_ticket_price ~ avg_ticket_price + seats_sold + show, 
                data = model_data,
                keep_reps = TRUE,
                model_fn = "lm")
top_model_intervals
```

```{r plotting top model confidence intervals, echo=FALSE}
top_model_intervals%>% 
  select(term, .replicates) %>% 
  unnest(cols = .replicates) %>% 
  ggplot(aes(x = estimate)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~ term, scales = "free_x") + 
  geom_vline(data = top_model_intervals, aes(xintercept = .lower), col = "red") + 
  geom_vline(data = top_model_intervals, aes(xintercept = .upper), col = "red") + 
  geom_vline(xintercept = 0, col = "green")
```


## Average Ticket Price Model

```{r avg_model, echo=FALSE, message=FALSE}
avg_model <- lm(formula = avg_ticket_price ~
                top_ticket_price +
                seats_sold + 
                show,  
             data = hamilton_and_lion_king)
summary(avg_model)

```

$$
Average Ticket Price = -40.78 +0.32 (top ticket price) + 0.007 (seats sold) -11.92 (show) + 27.49
$$

The results show that there is a statistically significant relationship between the top ticket price and seats sold at a show. Therefore the null hypothesis is rejected.

|  | Value | Interpretation |
|:-:|:-:|:-|
| R-Squared | 83.42% | The R-squared value shows that about 83.42% of the variation in the average ticket price can be explained by the avg model. |
| Intercept | -40.78 | We would expect an average of \$-40.78 for the average ticket price for The Lion King when the top ticket price is \$0 and 0 seats are sold. |
| Top ticket | 0.32 | While holding the other variables constant, for each \$1 difference in the top ticket price for The Lion King, an average of $0.32 difference is expected for average ticket price |
| Seats sold | 0.007 | While holding the other variables constant, for each seat sold for The Lion King, the average ticket price is expected to increase by an average of $0.007. |
| Hamilton show | -11.92 | Compared to The Lion King, we would expect that the average ticket price for Hamilton is $11.92 cheaper while holding other variables constant. |

### Plotting Average Model

```{r plotting avg_model, echo=FALSE, message=FALSE, warning=FALSE}
hamilton_and_lion_king %>%
  ggplot(aes(x = top_ticket_price, seats_sold, y = avg_ticket_price, color = show)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(x = "Top Ticket Price (USD) + Seats Sold",
       y = "Average Ticket Price (USD)",
       color = "Show") +
  theme_classic()
```


### Plotting Avg Model Residuals


```{r avg_model residuals, message=FALSE, include=FALSE}
broom::augment(avg_model)
```


```{r plotting avg_model residuals, echo=FALSE}
ggplot(avg_model, 
       aes(x = .resid)) + 
  geom_histogram(binwidth = 1) +
  theme_classic() +
  labs(y = "Number of occurrences",
       x = "Residuals")
```

### Avg Model Confidence Intervals

```{r avg model diagnostics, message=FALSE, include=FALSE}
broom::glance(avg_model)
```

```{r avg model coefficient estimates, message=FALSE, include=FALSE}
broom::tidy(avg_model)
```
```{r include=FALSE}
confint(avg_model)
```

```{r avg model confidence interval prep, include=FALSE}
set.seed(1407)

avg_model_intervals <- 
  reg_intervals(avg_ticket_price ~ top_ticket_price + seats_sold + show, 
                data = model_data,
                keep_reps = TRUE,
                model_fn = "lm")
avg_model_intervals
```


```{r plotting avg model confidence intervals, echo=FALSE}
avg_model_intervals%>% 
  select(term, .replicates) %>% 
  unnest(cols = .replicates) %>% 
  ggplot(aes(x = estimate)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~ term, scales = "free_x") + 
  geom_vline(data = avg_model_intervals, aes(xintercept = .lower), col = "red") + 
  geom_vline(data = avg_model_intervals, aes(xintercept = .upper), col = "red") + 
  geom_vline(xintercept = 0, col = "green")
```


\newpage

## Summary

```{r display model, echo=FALSE}
huxreg(top_model, avg_model)
```

\newpage

# Discussion {discussion}

## “Correlation does not imply causation”

Although comparing ticket prices between The Lion King and Hamilton quantified the price differences between the shows, this comparison was not useful in *predicting* ticket prices. For example, the model suggests that for each seat sold for The Lion King, the average ticket price is expected to increase by an average of \$0.007. Does that mean that if 2 seats are sold that the average ticket price will increase by \$0.014? Or if a 1,000 seats sold that the average ticket price will increase by $7? No. Although these variables are correlated, it is not a case of causality. 

There are other factors that affect ticket prices that this model did not capture. Take the season, for example. Holidays are known to attract more traffic, hence the number of tickets and seats sold is expected to increase.


## Weaknesses

The avg model's confidence interval for 'show - Hamilton' variable is not significant as a portion of the residuals were equal to zero. This could be caused by outlier data points. This also caused an error in the model's result that the average ticket price for Hamilton is $11.92 dollars cheaper The Lion King's. It should be investigated further to fix that error.


## Next Steps

It would be interesting to conduct a multiple regression model with different dependent variables to see how they correlate with ticket prices and number of seats sold, for example:

- Weekly grosses 

- Awards received, like Tony Awards

It would also be interesting to conduct an ARIMA model to forecast if and when Hamilton's grosses will exceed The Lion King's.

\newpage

\appendix

# Appendix {-}

```{r aveticketsvstopticketprice, echo=FALSE, fig.cap="Top ticket prices vursus average ticket prices", message=FALSE, warning=FALSE}
hamilton_and_lion_king %>% 
  ggplot(aes(x = avg_ticket_price, y = top_ticket_price)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Top Ticket Price (USD)") +
  xlab("Average Ticket Price (USD)") +
  theme_minimal()
```


```{r aveticketsvstopticketpricebyyear, echo=FALSE, fig.cap="Top ticket prices vursus average ticket prices per year", message=FALSE, warning=FALSE}
hamilton_and_lion_king %>% 
  mutate(year_of_show = lubridate::year(week_ending)) %>% 
  ggplot(aes(x = avg_ticket_price, y = top_ticket_price)) +
  geom_point(aes(color = show), alpha = 0.8) +
  ylab("Top Ticket Price (USD)") +
  xlab("Average Ticket Price (USD)") +
  theme_minimal() +
  facet_wrap(vars(year_of_show))
```


\newpage


# References


